---
title: "MYH7 promoter motifs"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
options(stringsAsFactors = FALSE)
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(biomartr)
library(biomaRt)
library(GenomicRanges)
```

## Retrieve MYH7 promoter motif matches from Ensembl

The Ensembl Regulatory Build seems to be one of the easiest / best resources for matched motifs in the human genome. I beleive that they also have information on bound proteins from ChIP-seq tracks -- which is awesome. Here we aim to do an analysis of the MYH7 promoter region to see which TFs could be linking MYH7b expression levels to MYH7 mRNA expression. 

```{r}
# Retrieve motif matches for MYH7

# Let's use the annotation we've been using
gtf <- rtracklayer::import("../../util/gencode.v33.annotation.gtf")
genes <- gtf[which(gtf$type == "gene")]

myh7 <- genes[which(genes$gene_name == "MYH7")]
myh7_promoter <- promoters(myh7)

biomartr::getMarts()
biomaRt::listMarts()

# Okay, so the biomart is unresponsive, so I'm downloading the JASPAR dataset from 
# UCSC. It's a bit big.. so I'm going to set this aside while it downloads.
# wget http://expdata.cmmt.ubc.ca/JASPAR/downloads/UCSC_tracks/2020/JASPAR2020_hg38.bb
```

